---
title: "EDA Joey"
author: "Joey Gilmartin"
date: "2026-02-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(dplyr, forcats)
```

## Data Visualizations
```{r}

## home / away success rate
home_away_success <- down4 |>
  group_by(posteam_type) |> 
  summarise(
    attempts = n(),
    success_rate = mean(fourth_down_converted, na.rm = TRUE)
  )

ggplot(home_away_success, 
       aes(x = posteam_type, 
           y = success_rate, 
           fill = posteam_type)) +
  geom_col(width = 0.6) +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Fourth Down Conversion Rate: Home vs Away",
    x = "Offensive Team Type",
    y = "Conversion Rate"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

## coach avg success rate (only take coaches with more than avg)
  ## maybe split on something
coach_success <- down4 |> 
  group_by(home_coach) |> 
  summarise(
    attempts = n(),
    conversion_rate = mean(fourth_down_converted, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  filter(attempts > 60)

ggplot(coach_success,
       aes(x = reorder(home_coach, conversion_rate),
           y = conversion_rate)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "4th Down Conversion Rate by Coach",
    x = "",
    y = "Conversion Rate"
  ) +
  theme_minimal()

## NOTES: Sean McVay is the greatest offensive mind ever

## drive play count / time of possession


## attempts at each yard marker
yard_summary <- down4 |> 
  group_by(yardline_100) |> 
  summarise(
    attempts = n(),
    conversion_rate = mean(fourth_down_converted, na.rm = TRUE),
    .groups = "drop"
  )
  ## colored by 
yard_summary |> filter(yardline_100 >5) |>
ggplot(aes(x = yardline_100,
           y = attempts,
           color = conversion_rate)) +
  geom_point(size = 3) +
  scale_x_continuous(
    breaks = seq(0, 100, 10),          # major breaks every 10
    minor_breaks = seq(0, 100, 5),     # minor ticks every 5
  ) +
  scale_color_viridis_c(labels = scales::percent_format()) +
  labs(
    title = "4th Down Attempts by Yardline",
    x = "Yards From Opponent Endzone",
    y = "Attempts",
    color = "Conversion Rate"
  ) +
  theme_minimal()


# # Using bins showed absolutely no trend  slight downsloped line across 60-55%
# yard_bins <- down4 |>
#   mutate(
#     yard_bin = cut(
#       yardline_100,
#       breaks = seq(0, 100, by = 5),
#       include.lowest = TRUE,
#       right = TRUE
#     )
#   ) |>
#   group_by(yard_bin) |>
#   summarise(
#     attempts = n(),
#     conversion_rate = mean(fourth_down_converted, na.rm = TRUE),
#     .groups = "drop"
#   )
# ggplot(yard_bins,
#        aes(x = yard_bin,
#            y = conversion_rate,
#            group = 1)) +
#   geom_line() +
#   geom_point() +
#   scale_y_continuous(labels = scales::percent_format()) +
#   labs(
#     title = "4th Down Conversion Rate by Field Position (5-Yard Bins)",
#     x = "Yards From Opponent Endzone",
#     y = "Conversion Rate"
#   ) +
#   theme_minimal() +
#   theme(
#     axis.text.x = element_text(angle = 90, vjust = 0.5)
#   )

down4 %>%
  mutate(
    yard_bin = floor(yardline_100 / 5) * 5,
    yd_bin = pmin(ydstogo, 10)  # cap long distances
  ) %>%
  group_by(yard_bin, yd_bin) %>%
  summarise(rate = mean(fourth_down_converted, na.rm = TRUE),
            attempts = n(),
            .groups = "drop") %>%
  ggplot(aes(x = yard_bin, y = yd_bin, fill = rate)) +
  geom_tile() +
  scale_fill_viridis_c(labels = scales::percent_format()) +
  labs(
    title = "4th Down Conversion Heatmap",
    x = "Yards From Opponent Endzone",
    y = "Yards To Go",
    fill = "Conversion Rate"
  ) +
  theme_minimal()

ggplot(down4,
       aes(x = score_differential,
           y = fourth_down_converted)) +
  geom_smooth(method = "loess", se = FALSE) +
  #scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Conversion Probability by Score Differential",
    x = "Score Differential (Offense - Defense)",
    y = "Conversion Probability"
  ) +
  theme_minimal()


down4 |> 
  group_by(drive_play_count) |> 
  summarise(
    attempts = n(),
    conversion_rate = mean(fourth_down_converted, na.rm = TRUE),
    .groups = "drop"
  ) |> 
ggplot(aes(x = drive_play_count,
           y = conversion_rate)) +
  geom_col() +
  #scale_y_continuous(labels = scales::percent_format()) +
  labs(
    title = "Conversion Probability by Drive Length",
    x = "Plays in Drive",
    y = "Conversion Probability"
  ) +
  theme_minimal()

```

```{r}
heatmap_data <- down4 %>%
  mutate(
    yard_bin = floor(yardline_100 / 5) * 5,
    ydstogo_bin = pmin(ydstogo, 10)
  ) %>%
  group_by(yard_bin, ydstogo_bin) %>%
  summarise(
    attempts = n(),
    conversion_rate = mean(fourth_down_converted, na.rm = TRUE),
    .groups = "drop"
  )


ggplot(heatmap_data,
       aes(x = yard_bin,
           y = ydstogo_bin)) +
  geom_tile(aes(fill = conversion_rate), color = "white") +
  geom_point(aes(size = attempts), alpha = 0.6) +
  scale_fill_viridis_c(labels = scales::percent_format()) +
  scale_size(range = c(1,6)) +
  scale_x_continuous(
    breaks = seq(0,100,10),
    minor_breaks = seq(0,100,5)
  ) +
  scale_y_continuous(breaks = 1:10) +
  labs(
    title = "4th Down Conversion Probability by Situation",
    subtitle = "Color = Conversion Rate | Size = Number of Attempts",
    x = "Yards From Opponent Endzone",
    y = "Yards To Go",
    fill = "Conversion Rate",
    size = "Attempts"
  ) +
  theme_minimal()
```


## Early Model Building - Answering Coach Question
```{r}
library(lme4)
library(dplyr)

# --- Prep -----------------------------------------------------------
down4_model <- down4 |>
  mutate(
    ydstogo    = factor(ydstogo),                          # 1, 2, 3 as categories
    play_type  = factor(play_type, levels = c("run", "pass")), # run as reference
    roof       = factor(roof),
    surface    = factor(surface),
    div_game   = factor(div_game),
    # Scale continuous predictors
    yardline_100_s        = scale(yardline_100),
    score_diff_s          = scale(score_differential),
    game_seconds_s        = scale(game_seconds_remaining),
    cum_conversion_rate_s = scale(cum_conversion_rate)
  ) |>
  filter(!is.na(fourth_down_converted))                   # drop any stray NAs


# --- Model 1: Core Situational --------------------------------------
m1 <- glm(
  fourth_down_converted ~
    ydstogo +
    yardline_100_s +
    score_diff_s +
    game_seconds_s,
  data   = down4_model,
  family = binomial
)

# --- Model 2: Add Play Design ---------------------------------------
m2 <- glm(
  fourth_down_converted ~
    ydstogo +
    yardline_100_s +
    score_diff_s +
    game_seconds_s +
    play_type +
    shotgun +
    no_huddle,
  data   = down4_model,
  family = binomial
)

# --- Model 3: Full Preliminary --------------------------------------
m3 <- glm(
  fourth_down_converted ~
    ydstogo +
    yardline_100_s +
    score_diff_s +
    game_seconds_s +
    play_type +
    shotgun +
    no_huddle +
    cum_conversion_rate_s +
    surface +
    roof +
    div_game,
  data   = down4_model,
  family = binomial
)

# --- Comparison -----------------------------------------------------
AIC(m1, m2, m3)

# Likelihood ratio tests (nested models)
anova(m1, m2, test = "Chisq")   # Does play design improve over situational alone?
anova(m2, m3, test = "Chisq")   # Does full set improve over play design?

# Summaries
summary(m1)
summary(m2)
summary(m3)


# how to unscale model terms
#0.110 / sd(down4_model$yardline_100)


# designed qb runs are a really good predictor
```

```{r}
m_final <- glm(
  fourth_down_converted ~
    ydstogo +
    yardline_100_s +
    play_type,
  data   = down4_model,
  family = binomial
)

summary(m_final)
AIC(m2, m_final)
```


# Mixed Effect Model
## Coaching 

```{r}
library(lme4)

m_mixed <- glmer(
  fourth_down_converted ~
    ydstogo +
    yardline_100_s +
    play_type +
    (1 | posteam_coach),
  data   = down4_model,
  family = binomial,
  control = glmerControl(optimizer = "bobyqa")
)

summary(m_mixed)

AIC(m_final, m_mixed)

```

```{r}
# Also # Attempt counts for the join
coach_counts <- down4_model |>
  group_by(posteam_coach) |>
  summarise(n = n())

# Extract random effects
coach_eff <- ranef(m_mixed) |>
  data.frame() |>
  select(posteam_coach = grp, condval)

coach_eff <- left_join(
  x = coach_eff,
  y = coach_counts,
  by = "posteam_coach"
)|>
  arrange(-condval) |>
  mutate(`odds of success` = exp(condval) |> round(2)) |>
  select(posteam_coach, attempts = n, `odds of success`, condval)

coach_eff

VarCorr(m_mixed)
```

## Plotting Coach Effects

```{r}
## wins? better teams convert more?
gg_coach_eff <- 
  coach_eff |> 
  # Only keeping rows with at least 100 attempts
  filter(attempts >= 80) |> 
  # blue = above average, red = below average and reordering the kickers
  mutate(
    bar_col = if_else(condval > 0, "#013369", "#D50A0A"),
    posteam_coach = fct_reorder(posteam_coach, condval)
  ) |> 
  
  # Bar chart for kicker effects
  ggplot(
    mapping = aes(
      x = forcats::fct_rev(posteam_coach),
      y = abs(condval),
      fill = bar_col
    )
  ) + 
  
  # Adding a line for the bars to sit on at 0
  geom_hline(
    linewidth = 1,
    yintercept = 0
  ) +
  
  geom_col(
    color = "black"
  ) +
  
  scale_fill_identity() + 
  
  scale_y_continuous(
    limits = c(0, 0.2),
    breaks = (0):5/10,
    expand = c(0,0),
    labels = round(exp(0:5/10), 1)
  ) + 
  
  labs(
    x = NULL,
    y = "Odd of success", 
    title = "Odds of 4th Down Conversion for NFL Head Coaches",
    subtitle = "2016 - 2025 regular season; Minumum 80 attempts"
  ) + 
  
  theme(
    axis.text.x = element_text(
      angle = 90,
      hjust = 0.95,
      vjust = 0.15
    )
  )



gg_coach_eff
```


## Coach Team Combination
```{r}
# Create coach-team combination column
down4_model <- down4_model |>
  mutate(coach_team = paste(posteam_coach, posteam, sep = "_"))

# Fit mixed model with coach-team random effect (no filtering)
m_mixed_coach_team <- glmer(
  fourth_down_converted ~
    ydstogo +
    yardline_100_s +
    score_diff_s +
    play_type +
    no_huddle +
    (1 | coach_team),
  data    = down4_model,
  family  = binomial,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(m_mixed_coach_team)

# Extract BLUPs
coach_team_ranef <- ranef(m_mixed_coach_team)$coach_team |>
  tibble::rownames_to_column("coach_team") |>
  rename(ranef = `(Intercept)`) |>
  arrange(desc(ranef))

print(coach_team_ranef, n = Inf)

VarCorr(m_mixed_coach_team)
```


## Coach team filtered
```{r}
# Attempts per coach-team combo
coach_team_attempts <- down4_model |>
  group_by(coach_team) |>
  summarise(attempts = n()) |>
  arrange(attempts)

print(coach_team_attempts)
mean(coach_team_attempts$attempts)

# Filter to at or above average
avg_attempts_ct <- mean(coach_team_attempts$attempts)

coach_team_keep <- coach_team_attempts |>
  filter(attempts >= avg_attempts_ct) |>
  pull(coach_team)

down4_filtered_ct <- down4_model |>
  filter(coach_team %in% coach_team_keep)

# Refit
m_mixed_coach_team_filtered <- glmer(
  fourth_down_converted ~
    ydstogo +
    yardline_100_s +
    score_diff_s +
    play_type +
    no_huddle +
    (1 | coach_team),
  data    = down4_filtered_ct,
  family  = binomial,
  control = glmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 2e5))
)

summary(m_mixed_coach_team_filtered)
VarCorr(m_mixed_coach_team_filtered)


# Coach-team attempt counts (already have this but need it for the join)
coach_team_counts <- down4_model |>
  group_by(coach_team) |>
  summarise(n = n())

# Extract random effects using your pattern
coach_team_eff <- ranef(m_mixed_coach_team_filtered) |>
  data.frame() |>
  select(coach_team = grp, condval)

coach_team_eff <- left_join(
  x = coach_team_eff,
  y = coach_team_counts,
  by = "coach_team"
)

# Top and bottom 5
bind_rows(
  .id = "type",
  "best"  = coach_team_eff |> slice_max(condval, n = 5),
  "worst" = coach_team_eff |> slice_min(condval, n = 5)
) |>
  arrange(-condval) |>
  mutate(`odds of success` = exp(condval) |> round(2)) |>
  select(coach_team, attempts = n, `odds of success`)
```


hard to get accurate 
- could try bootstrapping
- explore interpreting the standard deviation
  - why do coaches vary so much


# Momentum Analysis

Hold constant: Yards to go, yardline 100
Use drive play count, time of possession, drive first downs, 
  cumulative success rate

```{r}
momentum_mod <- glm(
  fourth_down_converted ~
    ydstogo +
    yardline_100 +
    drive_first_downs +
    posteam_winning + play_in_rz,
  data   = down4,
  family = binomial
)

summary(momentum_mod)


## notes
# cumulative success rate is not significant alone
# drive play count is significant- but is not when drive first down is added to the model
# same thing for time of possession
```



```{r}

```

